### 2007-01-12 Documentation/Paper2/paper2.tex
*** empty log message ***

\section{Simulation of non-standard physics}

The probability calculation in \GLOBES\ 
Eg. Damping, Hamiltonian, MAVANS, Toshi etc.

@@

\GLOBES\ supports four types of systematical errors as a standard: Signal and background
normalization errors, as well as signal and background tilt or energy calibration errors.
The tilt (T) is impemented as a linear distortion of the spectrum around the center,
and the energy calibration (C) as a distortion (stretching) of the reconstructed energy scale.
For the implementation of these systematical errors, the  ``pull method'' is used~\cite{Fogli:2002pt}, which introduces  so-called nuisance parameters $\zeta_i$. 
For example, for the signal and background normalization errors, the rates in
each bin $i$ are multiplied with
\begin{equation}
s_i(\zeta_1) := (1+\zeta_1)\cdot s_i \, , \quad b_i(\zeta_2) : =(1+\zeta_2)\cdot b_i \, ,
\end{equation} 
\ie, the rates are scaled by these parameters. The systematics $\chi^2$ is then minimized over
these parameters:
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( \,
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} \frac{\zeta_j^2}{\sigma_{\zeta_j}^2}
\, \right)\,.
\label{equ:chipull}
\end{equation}
Here $\chi^2(\boldsymbol{\lambda}, \zeta_1, \ldots, \zeta_k)$ is the usual Possonian $\chi^2$
depending on the neutrino oscillation parameters $\boldsymbol{\lambda}$ and the nuisance
parameters $\zeta_i$ (which, for instance, scale the signal or background rates).
In addition, Gaussian penalties $\zeta_j^2/\sigma_{\zeta_j}^2$ are added, where 
$\sigma_{\zeta_j}$ corresponds to the actual systematical error.
Core part of \equ{chipull} is the definition of $\chi^2$. For example, for a background-free measurement and
a signal normalization error only, it reads
\begin{equation}
  \chi^2 (\boldsymbol{\lambda},a) =  \sum_{i=1}^{\textrm{\# of bins}}2
     \left( O_{i} - (1 + a) T_{i} +
       (1 + a) T_{i} \log \frac{(1 + a) T_{i}}{O_{i}}  \right) + \frac{a^2}{\sigma_a^2} 
\label{equ:chistandard}
\end{equation}
(JK/PH: CHECK THIS EQUATION IF THIS IS THE ACTUAL IMPLEMENTATION. HERE THE SYSTEMATICS IS\\
IMPLEMENTED IN THE SIMULATED RATES. IN ADDITION, CHECK IF THEORETICAL AND EXP. RATES
ARE CORRECTLY GIVEN. [WW])

@@

Using \GLOBESN\ and higher, one can modify the standard $\chi^2$ function in
\GLOBES\ which corresponds to \equ{chistandard}. In addition, the systematics
concept of ``error dimensions'' is replaced by a concept directly related to
this $\chi^2$ function. As an example, consider a reactor experiment with near ($N$)
and far ($F$) detectors which are described by a reactor flux uncertainty $\sigma_R$
and two fiducial volume errors of the near $\sigma_N$ and far $\sigma_F$ detectors.
Then \equ{chistandard} will be replaced by
\begin{align}
  \chi^2(\boldsymbol{\lambda},a_N,a_F,a_R) = & \sum_{i=1}^{\textrm{\# of bins}} \sum_{d = N,F} 2
     \left( O_{d,i} - (1 + a_R + a_d) T_{d,i} + \right. \nonumber \\
& \left. + (1 + a_R + a_d) T_{d,i} \log \frac{(1 + a_R + a_d) T_{d,i}}{O_{d,i}}  \right) 
          + \frac{a_R^2}{\sigma_R^2} + \frac{a_N^2}{\sigma_N^2} + \frac{a_F^2}{\sigma_F^2}
\label{equ:chi2reactor}
\end{align}
HOW CAN ONE MAKE THE LEFT BRACKET LARGER? [WW]
This function can be defined with user-defined systenatics. Similarly, calibration errors, shape
errors, uncorrelated bin-to-bin errors, \etc , can be introduced. Furthermore, more complicated setups using multiple discrete sources and detectors can be simulated. For details, we refer to the manual~\cite{Manual}.
In the past, user-defined systematics has (as experimental feature) been used in \Ref~\cite{Huber:2006vr} to
simulate Double Chooz and in \Ref~\cite{Barger:2006kp} to simulate T2KK. ANYTHING ELSE? [WW]

@@

>>>>

\section{Oscillation probabilities and the simulation of non-standard physics}

The probability calculation in \GLOBES\ is based on the diagonalization of the
Hamiltonian in layers of constant matter density using the standard three flavor
scenario of neutrino oscillations. In flavor base, we have
\begin{equation}
\mathcal{H}(n_e) = \frac{1}{2E} \, U \,
\left(
\begin{array}{ccc}
0 & 0 & 0 \\
0  & \sdm & 0 \\
0  & 0 & \ldm \\
\end{array}
\right) \,
U^\dagger + 
\left(
\begin{array}{ccc}
 \pm \sqrt{2} G_F n_e & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & 0 
\end{array}
\right)
\label{equ:ham}
\end{equation}
with $U$ the mixing matrix (described by four parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, and $\deltacp$), and $n_e$ the constant electron density
in that matter density layer. The electron density is related to the matter density $\rho$ by 
$n_e \simeq 0.5 \, \rho/m_N$ with $m_N$ being the nucleon mass. The sign in the second term depends on if one
uses neutrinos or antineutrinos. For $k$ matter density layers with densities
$n^k_e$ and thicknesses $x_i$, the oscillation probability then evaluates to
\begin{equation}
P(\nu_\alpha \rightarrow \nu_\beta) = \left| \langle \nu_\beta | \mathcal{E}(n^k_e, x_k) \hdots  \mathcal{E}(n^1_e, x_1) | \nu_\alpha \rangle \right|^2
\label{equ:probs}
\end{equation}
with the evolution operators (CORRECT REFERENCE?)
\begin{equation}
  \mathcal{E}(n^k_e, x_k) = \exp \left( -i \mathcal{H}(n^k_e) x_k \right) = \frac{1}{2E} \, \hat{U}(n^k_e) \, \left( \begin{array}{ccc}
e^{-i \, m_1^k \, x_k} & 0 & 0 \\
0 &  e^{-i \, m_2^k \, x_k} & 0 \\
0 & 0 & e^{-i \, m_3^k \, x_k} 
\end{array}
\right) \, \hat{U}(n^k_e)^\dagger \, .
\end{equation}
Here the Hamiltonian $\mathcal{H}(n^k_e)$ is diagonalized by the unitary mixing matrix in matter $\hat{U}(n^k_e)$
with the eigenvalues $m_1^k$, $m_2^k$, and $m_3^k$. The probabilities in \equ{probs} are used for the
event rate compution, which is described in greater detail in \Ref~\cite{Huber:2004ka}. In the next section,
we will demonstrate how systematics is implemented thereafter.

\GLOBESN\ allows the modification of this Hamiltonian and the whole probability engine without
re-compilation of the \GLOBES\ software. This is implemented
by using pointers for the functions calling the standard probability engine, which can be changed if
a different probability engine is registered. For example, a user may just copy the standard probability
engine of \GLOBES , and modify it, and register it. An important prerequisite for the ability to handle
more than the standard six oscillation parameters (plus matter density) $\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho)$, \ie,
\begin{equation}
\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho, \eta_1, \hdots, \eta_n)
\end{equation}
with $n$ non-standard parameters $\eta_j$. These parameters can be accessed in the usual way, see \Ref~\cite{Manual}.
For example, non-standard Hamiltonian effects in the $e$-$\tau$-sector may be introduced by modifying \equ{ham} to
\begin{equation}
 \mathcal{\hat{H}} = \mathcal{H}(n_e) + 
\left(
\begin{array}{ccc}
0 & 0 & \epsilon_{e \tau} \\
0 & 0 & 0 \\
\epsilon_{e \tau}^* & 0 & 0 
\end{array}
\right)
\end{equation}
In this case, there are two more real parameters, such as the absolute value and phase of $\epsilon_{e \tau}$.
In the literature, the non-standard physics feature in \GLOBES\ as experimental feature has been used for damping effects (such as neutrino decoherence, decay, \etc) in \Ref~\cite{Blennow:2005yk}, Hamiltonian-level effects (such as non-standard matter effects) in \Ref~\cite{Blennow:2005qj}, and mass-varying neutrinos in \Ref~\cite{Schwetz:2005fy} by implementing an environment dependence of neutrino mass.  ANY OTHERS (PH!)?

@@

\GLOBES\ supports four types of systematical errors as a standard: Signal and background
normalization errors, as well as signal and background tilt or energy calibration errors.
The tilt (T) is impemented as a linear distortion of the spectrum around the center,
and the energy calibration (C) as a distortion (stretching) of the reconstructed energy scale.
For the implementation of these systematical errors, the  ``pull method'' is used~\cite{Fogli:2002pt}, which introduces  so-called nuisance parameters $\zeta_i$. 
For example, for the signal and background normalization errors, the respective event rates 
$s_i$ and $b_i$ in
each bin $i$ are multiplied with\footnote{For details on the event rate calculation, see \Ref~\cite{Huber:2004ka}.}
\begin{equation}
s_i(\zeta_1) := (1+\zeta_1)\cdot s_i \, , \quad b_i(\zeta_2) : =(1+\zeta_2)\cdot b_i \, ,
\end{equation} 
\ie, the rates are scaled by these parameters. The systematics $\chi^2$ is then minimized over
these parameters:
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( \,
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} \frac{\zeta_j^2}{\sigma_{\zeta_j}^2}
\, \right)\,.
\label{equ:chipull}
\end{equation}
Here $\chi^2(\boldsymbol{\lambda}, \zeta_1, \ldots, \zeta_k)$ is the usual Possonian $\chi^2$
depending on the neutrino oscillation parameters $\boldsymbol{\lambda}$ and the nuisance
parameters $\zeta_i$ (which, for instance, scale the signal or background rates).
In addition, Gaussian penalties $\zeta_j^2/\sigma_{\zeta_j}^2$ are added, where 
$\sigma_{\zeta_j}$ corresponds to the actual systematical error.
Core part of \equ{chipull} is the definition of $\chi^2$. For example, for a background-free measurement and
a signal normalization error only, it reads
\begin{equation}
  \chi^2 (\boldsymbol{\lambda},a) =  \sum_{i=1}^{\textrm{\# of bins}}2
     \left( O_{i} - (1 + a) T_{i} +
       (1 + a) T_{i} \log \frac{(1 + a) T_{i}}{O_{i}}  \right) + \frac{a^2}{\sigma_a^2} 
\label{equ:chistandard}
\end{equation}
(JK/PH: CHECK THIS EQUATION IF THIS IS THE ACTUAL IMPLEMENTATION. HERE THE SYSTEMATICS IS\\
IMPLEMENTED IN THE SIMULATED RATES. IN ADDITION, CHECK IF THEORETICAL AND EXP. RATES
ARE CORRECTLY GIVEN. [WW])

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=9cm]{reactor}
\end{center}
\mycaption{\label{fig:reactor} Luminosity scaling of the $\stheta$ sensitivity for different systematical error assumptions (90\% confidence level). The thick curve corresponds to \equ{chi2reactor} for $\sigma_R=\sigma_{\mathrm{corr}}=2.8\%$, $\sigma_N=\sigma_F=\sigma_{\mathrm{uncorr}}=0.6\%$ and an additional calibration error. Figure similar to \Ref~\cite{Huber:2006vr}. MISLEADING: WHY IS THE LUMINOSITY
OF THE ``FAR DETECTOR'' ON THE X-AXIS? SHOULD THIS NOT BE THE SAME AS THE ONE FOR THE NEAR DETECTOR? 
AT LEAST IF I INTERPRET THE CODE CORRECTLY [WW]}
\end{figure}

Using \GLOBESN\ and higher, one can modify the standard $\chi^2$ function in
\GLOBES\ which corresponds to \equ{chistandard}. In addition, the systematics
concept of ``error dimensions'' is replaced by a concept directly related to
this $\chi^2$ function. As an example, consider a reactor experiment with near ($N$)
and far ($F$) detectors which are described by a reactor flux uncertainty $\sigma_R$
and two fiducial volume errors of the near $\sigma_N$ and far $\sigma_F$ detectors.
Then \equ{chistandard} will be replaced by
\begin{align}
  \chi^2(\boldsymbol{\lambda},a_N,a_F,a_R) = & \sum_{i=1}^{\textrm{\# of bins}} \sum_{d = N,F} 2
     \left( O_{d,i} - (1 + a_R + a_d) T_{d,i} + \right. \nonumber \\
& \left. + (1 + a_R + a_d) T_{d,i} \log \frac{(1 + a_R + a_d) T_{d,i}}{O_{d,i}}  \right) 
          + \frac{a_R^2}{\sigma_R^2} + \frac{a_N^2}{\sigma_N^2} + \frac{a_F^2}{\sigma_F^2}
\label{equ:chi2reactor}
\end{align}
HOW CAN ONE MAKE THE LEFT BRACKET LARGER? [WW]
This function can be defined with user-defined systematics. Similarly, calibration errors, shape
errors, uncorrelated bin-to-bin errors, \etc , can be introduced. We show in \figu{reactor} the result of
such a calculation, where the thick curve corresponds to \equ{chi2reactor} plus an additional 
energy calibration error.
Furthermore, more complicated setups using multiple discrete sources and detectors can be simulated. For details, we refer to the manual~\cite{Manual}.
In the past, user-defined systematics has (as experimental feature) been used in \Ref~\cite{Huber:2006vr} to
simulate Double Chooz and in \Ref~\cite{Barger:2006kp} to simulate T2KK. ANYTHING ELSE? [WW]

@@

TABLE FROM MANUAL?
